---
title: "Phase 3: Statistical Analysis — Telco Customer Churn"
author: "Muhammad Humza Majeed"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: cosmo
    highlight: tango
    self_contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(tidyverse)
library(rstatix)
library(ggpubr)
library(infer)
library(knitr)
library(kableExtra)
```
## 1. Load Data

```{r load-data}
df <- read.csv("../data/cleaned/telco_churn_cleaned.csv")

# Quick look
glimpse(df)
cat("Rows:", nrow(df), "| Columns:", ncol(df))
```

## 2. Descriptive Statistics by Churn Group

```{r descriptive-stats}
df %>%
  group_by(Churn) %>%
  summarise(
    Count         = n(),
    Avg_Tenure    = round(mean(tenure), 2),
    Avg_Monthly   = round(mean(MonthlyCharges), 2),
    Avg_Total     = round(mean(TotalCharges), 2),
    SD_Monthly    = round(sd(MonthlyCharges), 2)
  ) %>%
  kable(caption = "Descriptive Statistics by Churn") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## 3. Hypothesis Test 1: T-Test — Tenure by Churn

**H₀:** Mean tenure is the same for churned and non-churned customers  
**H₁:** Churned customers have significantly lower tenure

```{r ttest-tenure}
# Run the t-test
t_result <- df %>%
  t_test(tenure ~ Churn, alternative = "two.sided") %>%
  add_significance()

print(t_result)

# Visualize
ggboxplot(df, x = "Churn", y = "tenure",
          color = "Churn", palette = c("steelblue", "salmon"),
          add = "mean_ci",
          title = "Tenure by Churn Status",
          xlab = "Churn", ylab = "Tenure (months)") +
  stat_compare_means(method = "t.test", label.y = 75)
```

**Interpretation:** If p-value < 0.05, we reject H₀ — churned customers 
have significantly different tenure.

## 4. Hypothesis Test 2: T-Test — Monthly Charges by Churn

**H₀:** Mean MonthlyCharges is the same for both groups  
**H₁:** Churned customers pay significantly higher monthly charges

```{r ttest-monthly}
t_result2 <- df %>%
  t_test(MonthlyCharges ~ Churn, alternative = "two.sided") %>%
  add_significance()

print(t_result2)

# Visualize
ggboxplot(df, x = "Churn", y = "MonthlyCharges",
          color = "Churn", palette = c("steelblue", "salmon"),
          add = "mean_ci",
          title = "Monthly Charges by Churn Status",
          xlab = "Churn", ylab = "Monthly Charges ($)") +
  stat_compare_means(method = "t.test", label.y = 120)
```
## 5. Hypothesis Test 3: Chi-Square — Churn vs Contract Type

**H₀:** Churn is independent of Contract type  
**H₁:** There is a significant association between Churn and Contract type

```{r chisq-contract}
# Contingency table
contract_table <- table(df$Contract, df$Churn)
print(contract_table)

# Chi-square test
chisq_result <- chisq.test(contract_table)
print(chisq_result)

# Visualize
df %>%
  group_by(Contract, Churn) %>%
  summarise(Count = n(), .groups = "drop") %>%
  group_by(Contract) %>%
  mutate(Percentage = Count / sum(Count) * 100) %>%
  filter(Churn == "Yes") %>%
  ggplot(aes(x = Contract, y = Percentage, fill = Contract)) +
  geom_col(show.legend = FALSE) +
  scale_fill_manual(values = c("coral", "steelblue", "seagreen")) +
  labs(title = "Churn Rate by Contract Type",
       x = "Contract Type", y = "Churn Rate (%)") +
  theme_minimal()
```

**Interpretation:** If p-value < 0.05, contract type significantly 
affects whether a customer churns.

## 6. Hypothesis Test 4: Chi-Square — Churn vs Payment Method

```{r chisq-payment}
payment_table <- table(df$PaymentMethod, df$Churn)
chisq_payment <- chisq.test(payment_table)
print(chisq_payment)

# Visualize
df %>%
  group_by(PaymentMethod, Churn) %>%
  summarise(Count = n(), .groups = "drop") %>%
  group_by(PaymentMethod) %>%
  mutate(Percentage = Count / sum(Count) * 100) %>%
  filter(Churn == "Yes") %>%
  ggplot(aes(x = reorder(PaymentMethod, -Percentage), 
             y = Percentage, fill = PaymentMethod)) +
  geom_col(show.legend = FALSE) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Churn Rate by Payment Method",
       x = "Payment Method", y = "Churn Rate (%)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 15, hjust = 1))
```

## 7. Confidence Intervals — Monthly Charges by Churn Group

```{r confidence-intervals}
ci_results <- df %>%
  group_by(Churn) %>%
  summarise(
    Mean  = mean(MonthlyCharges),
    SE    = sd(MonthlyCharges) / sqrt(n()),
    Lower = Mean - 1.96 * SE,
    Upper = Mean + 1.96 * SE
  )

print(ci_results)

# Visualize CI
ggplot(ci_results, aes(x = Churn, y = Mean, color = Churn)) +
  geom_point(size = 4) +
  geom_errorbar(aes(ymin = Lower, ymax = Upper), width = 0.2, size = 1.2) +
  scale_color_manual(values = c("steelblue", "salmon")) +
  labs(title = "95% Confidence Intervals for Monthly Charges",
       x = "Churn", y = "Monthly Charges ($)") +
  theme_minimal()
```

## 8. Statistical Test Summary

```{r summary-table}
# T-test for Tenure
t_result <- t.test(tenure ~ Churn, data = df)

# T-test for Monthly Charges
t_result2 <- t.test(MonthlyCharges ~ Churn, data = df)

# Chi-Square tests
chisq_result <- chisq.test(table(df$Churn, df$Contract))
chisq_payment <- chisq.test(table(df$Churn, df$PaymentMethod))
## 8. Statistical Test Summary
# Create the data frame with corrected p-value references
summary_df <- data.frame(
  Test = c(
    "T-Test: Tenure vs Churn",
    "T-Test: MonthlyCharges vs Churn",
    "Chi-Square: Churn vs Contract",
    "Chi-Square: Churn vs PaymentMethod"
  ),
  Method = c("Independent t-test", "Independent t-test",
             "Chi-Square", "Chi-Square"),
  P_Value = c(
    format.pval(t_result$p.value, eps = .0001, digits = 4),
    format.pval(t_result2$p.value, eps = .0001, digits = 4),
    format.pval(chisq_result$p.value, eps = .0001, digits = 4),
    format.pval(chisq_payment$p.value, eps = .0001, digits = 4)
  ),
  Significant = ifelse(
    c(t_result$p.value, t_result2$p.value, 
      chisq_result$p.value, chisq_payment$p.value) < 0.05,
    "Yes ✅", "No ❌"
  )
)

# Render the table
library(knitr)
library(kableExtra)

kable(summary_df, caption = "Summary of Hypothesis Tests") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "bordered"))
```
