---
title: "Phase 6: Advanced Visualizations — Telco Customer Churn"
author: "Muhammad Humza Majeed"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
    toc_depth: 2
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.path = "figures/",          # ← ADD THIS
  fig.align = "center"
)    
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE,
                      message = FALSE, fig.align = "center")

library(tidyverse)
library(plotly)
library(scales)
library(patchwork)
library(knitr)
library(kableExtra)
library(RColorBrewer)
```

## 1. Load Data

```{r load-data}
df <- read.csv("../data/cleaned/telco_churn_cleaned.csv")

# Quick summary
cat("Rows:", nrow(df), "| Columns:", ncol(df))
cat("\nChurn Rate:", round(mean(df$Churn == "Yes") * 100, 2), "%")
```

## 2. Churn Rate by Contract Type

```{r churn-contract}
contract_churn <- df %>%
  group_by(Contract, Churn) %>%
  summarise(Count = n(), .groups = "drop") %>%
  group_by(Contract) %>%
  mutate(Percentage = round(Count / sum(Count) * 100, 1))

p1 <- ggplot(contract_churn,
             aes(x = Churn, y = Percentage, fill = Churn)) +
  geom_col(width = 0.6, show.legend = FALSE) +
  geom_text(aes(label = paste0(Percentage, "%")),
            vjust = -0.5, size = 3.5, fontface = "bold") +
  facet_wrap(~ Contract) +
  scale_fill_manual(values = c("No" = "steelblue", "Yes" = "salmon")) +
  labs(title = "Churn Rate by Contract Type",
       subtitle = "Month-to-month contracts have the highest churn",
       x = "Churn", y = "Percentage (%)") +
  theme_minimal(base_size = 12) +
  theme(strip.text = element_text(face = "bold"),
        plot.title = element_text(face = "bold", size = 14))

p1
ggsave("../outputs/figures/churn_by_contract.png", p1, dpi = 150,
       width = 10, height = 5)
```

## 3. Monthly Charges Distribution

```{r monthly-dist}
p2 <- ggplot(df, aes(x = MonthlyCharges, fill = Churn)) +
  geom_density(alpha = 0.6) +
  scale_fill_manual(values = c("No" = "steelblue", "Yes" = "salmon")) +
  geom_vline(data = df %>% group_by(Churn) %>%
               summarise(mean_charge = mean(MonthlyCharges)),
             aes(xintercept = mean_charge, color = Churn),
             linetype = "dashed", linewidth = 1, show.legend = FALSE) +
  scale_color_manual(values = c("No" = "darkblue", "Yes" = "darkred")) +
  labs(title = "Monthly Charges Distribution by Churn",
       subtitle = "Churned customers tend to have higher monthly charges",
       x = "Monthly Charges ($)", y = "Density", fill = "Churn") +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold", size = 14))

p2
ggsave("../outputs/figures/monthly_charges_density.png", p2,
       dpi = 150, width = 10, height = 5)

```
## 4. Tenure vs Monthly Charges (Interactive)

```{r scatter-interactive}
p3 <- ggplot(df, aes(x = tenure, y = MonthlyCharges, color = Churn)) +
  geom_point(alpha = 0.4, size = 1.5) +
  scale_color_manual(values = c("No" = "steelblue", "Yes" = "salmon")) +
  labs(title = "Tenure vs Monthly Charges by Churn",
       x = "Tenure (months)", y = "Monthly Charges ($)") +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))

p3
ggsave("../outputs/figures/tenure_vs_monthly.png", p3,
       dpi = 150, width = 10, height = 5)
```

p3 <- ggplot(df, aes(x = tenure, y = MonthlyCharges,
                      color = Churn,
                      text = paste("Tenure:", tenure,
                                   "<br>Monthly:", MonthlyCharges,
                                   "<br>Contract:", Contract))) +
  geom_point(alpha = 0.4, size = 1.5) +
  scale_color_manual(values = c("No" = "steelblue", "Yes" = "salmon")) +
  labs(title = "Tenure vs Monthly Charges",
       x = "Tenure (months)", y = "Monthly Charges ($)") +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))

# Make it interactive with plotly
ggplotly(p3, tooltip = "text") %>%
  layout(title = list(text = "Tenure vs Monthly Charges",
                      font = list(size = 14)))

> 💡 Hover over points to see customer details interactively.



## 5. Churn Heatmap

```{r churn-heatmap}
heatmap_df <- df %>%
  group_by(InternetService, Contract) %>%
  summarise(ChurnRate = round(mean(Churn == "Yes") * 100, 1),
            .groups = "drop")

p4 <- ggplot(heatmap_df,
             aes(x = Contract, y = InternetService, fill = ChurnRate)) +
  geom_tile(color = "white", linewidth = 1) +
  geom_text(aes(label = paste0(ChurnRate, "%")),
            size = 5, fontface = "bold", color = "white") +
  scale_fill_gradient(low = "steelblue", high = "darkred",
                      name = "Churn Rate (%)") +
  labs(title = "Churn Rate Heatmap",
       subtitle = "By Internet Service Type and Contract",
       x = "Contract Type", y = "Internet Service") +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold", size = 14),
        axis.text.x = element_text(angle = 15, hjust = 1))

p4
ggsave("../outputs/figures/churn_heatmap.png", p4,
       dpi = 150, width = 9, height = 5)
```

## 6. Tenure Distribution by Churn

```{r tenure-violin}
p5 <- ggplot(df, aes(x = Churn, y = tenure, fill = Churn)) +
  geom_violin(alpha = 0.6, trim = FALSE) +
  geom_boxplot(width = 0.15, fill = "white",
               outlier.shape = NA, coef = 0) +
  scale_fill_manual(values = c("No" = "steelblue", "Yes" = "salmon")) +
  labs(title = "Tenure Distribution by Churn",
       subtitle = "Churned customers leave much earlier",
       x = "Churn", y = "Tenure (months)") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none",
        plot.title = element_text(face = "bold", size = 14))

p5
ggsave("../outputs/figures/tenure_violin.png", p5,
       dpi = 150, width = 8, height = 5)
```

## 7. Churn Rate Across Services

```{r lollipop}
service_cols <- c("PhoneService", "MultipleLines", "OnlineSecurity",
                  "OnlineBackup", "DeviceProtection", "TechSupport",
                  "StreamingTV", "StreamingMovies")

service_churn <- sapply(service_cols, function(col) {
  mean(df$Churn[df[[col]] == "Yes"] == "Yes", na.rm = TRUE) * 100
})

lollipop_df <- data.frame(
  Service   = service_cols,
  ChurnRate = round(service_churn, 1)
) %>% arrange(ChurnRate)

lollipop_df$Service <- factor(lollipop_df$Service,
                               levels = lollipop_df$Service)

p6 <- ggplot(lollipop_df, aes(x = Service, y = ChurnRate)) +
  geom_segment(aes(xend = Service, y = 0, yend = ChurnRate),
               color = "steelblue", linewidth = 1.2) +
  geom_point(size = 5, color = "salmon") +
  geom_text(aes(label = paste0(ChurnRate, "%")),
            hjust = -0.4, size = 3.5, fontface = "bold") +
  coord_flip() +
  ylim(0, max(lollipop_df$ChurnRate) + 10) +
  labs(title = "Churn Rate by Service Subscription",
       subtitle = "Customers without online security churn the most",
       x = NULL, y = "Churn Rate (%)") +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold", size = 14))

p6
ggsave("../outputs/figures/service_churn_lollipop.png", p6,
       dpi = 150, width = 9, height = 6)
```

## 8. Combined Dashboard

```{r dashboard, fig.width=14, fig.height=10}
dashboard <- (p1 | p2) / (p5 | p4) +
  plot_annotation(
    title    = "Customer Churn Analytics Dashboard",
    subtitle = "Telco Customer Churn — End-to-End Analysis",
    theme    = theme(
      plot.title    = element_text(size = 18, face = "bold", hjust = 0.5),
      plot.subtitle = element_text(size = 12, hjust = 0.5, color = "gray40")
    )
  )

dashboard
ggsave("../outputs/figures/final_dashboard.png", dashboard,
       dpi = 150, width = 14, height = 10)
```

## 9. Key Business Insights

```{r insights-table}
insights <- data.frame(
  Finding = c(
    "Month-to-month churn rate",
    "Two-year contract churn rate",
    "Fiber optic customers churn rate",
    "Customers without online security",
    "High monthly charges (>$65) churn rate",
    "Low tenure (<12 months) churn rate"
  ),
  Value = c("~42%", "~3%", "~42%", "~45%",
            "~40%", "~50%"),
  Business_Action = c(
    "Offer discounts to switch to annual plans",
    "Prioritize long-term contract incentives",
    "Improve fiber service quality & pricing",
    "Bundle online security into basic plans",
    "Review high-charge pricing strategy",
    "Improve onboarding for new customers"
  )
)

kable(insights,
      col.names = c("Finding", "Churn Rate", "Recommended Action"),
      caption   = "Key Insights & Business Recommendations") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "bordered")) %>%
  column_spec(3, bold = TRUE, color = "steelblue")
```

```


```